---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# visR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of visR is to enable fit-for-purpose, reproducible and reusable clinical and medcial research focused visualizations and tables with sensible defaults and based on sound graphical principles.

## Motivation
By using a common package for visualising data analysis results in the clinical development process, we want to have a **positive influence** on 

* **choice of visualisation** by making it easy explore different visualisation and to use impactful visualisations fit-for-purpose
* effective visual communication by making it easy to **implement best practices** 

We are not judging on what visualisation you chose for your research question, but want facilitate to make you do your work as you need it!

You can read more about the philosophy and architecture in the repo wiki.

## Installation

Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openpharma/visR")
```

## Example

This is a basic example which shows you how to solve a common problem:

TODO: add an example. 

```{r example}
library(visR)


```

Please note that the 'visR' project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.
