---
title: "Example Survival analysis using visR"
author: Charlotta Fruechtenicht
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# Not needed
#library(visR)
```


# Time to Event Analysis using visR

This short tutorial illustrates a typical use case in clinical development - the analysis of time to a certain event (e.g., death) in different populations. Typically, data obtained in randomized clinical trials (RCT) can be used to estimate the overall survival of patients in one group (e.g., treated with drug X) vs another group (e.g., treated with drug Y) and thus determine if there is a difference between these treatments.

For a more thorough introduction to Survival Analysis, we recommend the following tutorial: https://bioconnector.github.io/workshops/r-survival.html

In this example, we will work with patient data from The Cancer Genome Atlas and compare survival in primary breast cancer who received either chemotherapy or hormone therapy.

```{r installRTCGA, eval=FALSE}
#install.packages("BiocManager")
#BiocManager::install()
#BiocManager::install(c("RTCGA", "RTCGA.clinical"))
```
```{r installvisR, eval=FALSE}
#devtools::install_github("https://github.com/openpharma/visR.git")
```


```{r imports, echo=FALSE}
library(dplyr)
library(ggplot2)
library(RTCGA)
library(RTCGA.clinical)
library(visR)
```

```{r globalSetup}
DATASET <- paste0("TCGA BRCA dataset (RTCGA version ", packageVersion("RTCGA"), ")")

```



```{r initialCohort}
# names(BRCA.clinical)[grepl("ecog", names(BRCA.clinical) )]
# 
# head(BRCA.clinical, 50) %>% 
#   select(patient.bcr_patient_barcode, patient.distant_metastasis_present_ind2)

brca_cohort <- survivalTCGA(BRCA.clinical, 
                     extract.cols=c("admin.disease_code", 
                                    "patient.breast_carcinoma_estrogen_receptor_status",
                                    "patient.breast_carcinoma_progesterone_receptor_status",
                                    "patient.clinical_cqcf.tumor_type",
                                    "patient.drugs.drug.therapy_types.therapy_type",
                                    "patient.age_at_initial_pathologic_diagnosis")) %>% 
  rename(tumor_type = patient.clinical_cqcf.tumor_type,
         therapy = patient.drugs.drug.therapy_types.therapy_type,
         er_status = patient.breast_carcinoma_estrogen_receptor_status,
         progesterone_status = patient.breast_carcinoma_progesterone_receptor_status,
         dx_age = patient.age_at_initial_pathologic_diagnosis,
         followup_time = times
         ) %>% 
  mutate(er_status = factor(er_status),
         progesterone_status = factor(progesterone_status),
         therapy = factor(therapy),
         dx_age = as.numeric(dx_age),
         dx_age_group = factor(case_when(dx_age < 30 ~ "< 30y",
                                         dx_age >= 30 & dx_age <= 50 ~ "30-50y",
                                         dx_age > 50 & dx_age <= 70 ~ "51-70y",
                                         dx_age > 70 ~ "> 70y"),
                               levels=c("< 30y", "30-50y", "51-70y", "> 70y")))

head(brca_cohort)

```

# Cohort Selection
For this exercise we will only include patients with (1) primary breast cancer who have (2) received either chemotherapy or hormone therapy in our analysis
```{r fig.cap='Attrition Diagram of BRCA cohort'}
figures_directory = "../figures"

# original cohort
allN <- n_distinct(brca_cohort$bcr_patient_barcode)

# remove non-primary
brca_cohort <- brca_cohort %>% 
  filter(tumor_type == "primary")
f1N <- n_distinct(brca_cohort$bcr_patient_barcode)

# remove non-missing therapy
brca_cohort <- brca_cohort %>% 
  filter(!is.na(therapy))
f2N <- n_distinct(brca_cohort$bcr_patient_barcode)

# remove everything except chemo and hormone therapy
brca_cohort <- brca_cohort %>% 
  filter(therapy %in% c("chemotherapy", "hormone therapy"))
f3N <- n_distinct(brca_cohort$bcr_patient_barcode)

# remove pts without vital status
brca_cohort <- brca_cohort %>% 
  filter(!is.na(patient.vital_status))
f4N <- n_distinct(brca_cohort$bcr_patient_barcode)

# remove pts without vital status
brca_cohort <- brca_cohort %>% 
  filter(followup_time >= 0)
f5N <- n_distinct(brca_cohort$bcr_patient_barcode)

# create array of Ns
N_array = c(allN, f1N, f2N, f3N, f4N, f5N)
descriptions <- c(                                                              
    "All patiensts",                                                            
    "Primary disease",                                         
    "Non-missing therapy",                                         
    "Chemotherapy or hormone therapy",
    "Vital Status",
    "Positive follow up time"
)                                                                               
complement_descriptions <- c(                                                   
    "Having exclusion criterion: non-primary disease",                                            
    "Having exclusion criterion: missing therapy data",                                            
    "Having exclusion criterion: non chemo and non hormone therapy",
    "Having exclusion criterion: missing vital status",
    "Having exclusion criterion: negative follow up time"
)

#TODO: Issue with diagrammer to svg - CHECK
# Create attrition flowchart
#vr_attrition(N_array, descriptions, complement_descriptions, 
#    output_path = sprintf("%s/attrition_diagram.svg", figures_directory))

# Plot the attrition diagram
#knitr::include_graphics("../figures/attrition_diagram.svg")
```

# Cohort Overview
Prior to starting any survival analysis, we want to further inspect our cohort and see if the there are systematic differences between the chemotherapy and hormone therapy arms.
```{r}
brca_cohort_tab1 <- brca_cohort %>% 
  select(-bcr_patient_barcode, admin.disease_code) %>% 
  mutate(patient.vital_status = case_when(patient.vital_status == 0 ~ "alive",
                                          patient.vital_status == 1 ~ "dead",
                                          is.na(patient.vital_status) ~ "missing"))

# create table
tab1_groups <- vr_create_tableone(brca_cohort, groupCols = c("therapy"))

# render table
#tab1_groups %>% vr_render_tableone(caption = "BRCA patients")
```

# Overall Survival
## Survival Object
```{r}

brca_cohort_os <- brca_cohort %>% 
  rename(time = followup_time,
         status = patient.vital_status,
         trt = therapy)

os_obj <- vr_est_kaplan_meier(brca_cohort_os, "survival::Surv(time, status) ~ trt")


knitr::kable(os_obj) %>% 
    kableExtra::kable_styling()

```

## Kaplan-Meier 
```{r fig.cap="Comparison of survival in breast cancer patients having under hormone therapy vs. chemotherapy", fig.height=6, fig.width=6}
library(ggplot2)
library(ggsci)
library(pammtools)
library(ggpubr)

##TODO: check function ->  "unused arguments (time_unit = time_unit, data_source = data_source)"
#vr_kaplan_meier(
#    brca_cohort_os, 
#    data_source = DATASET, 
#    time_unit = "days",
#    title = "Comparison of survival in breast cancer patients having \nunder hormone therapy vs. #chemotherapy"
#)
```
Note: it would be good in the vr function to be able to provide user-defined names for time, event and group(s)
Also: stratification by more than one variable

Note2: there seems to be a bug in plot_risk_table where it expects the time_unit but is not provided time_unit in the function parameters
