---
title: "Survival Analysis with visR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis with visR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This short tutorial illustrates a typical use case in clinical development - the analysis of time to a certain event (e.g., death) in different populations. Typically, data obtained in randomized clinical trials (RCT) can be used to estimate the overall survival of patients in one group (e.g., treated with drug X) vs another group (e.g., treated with drug Y) and thus determine if there is a difference between these treatments.

For a more thorough introduction to Survival Analysis, we recommend the following tutorial: https://bioconnector.github.io/workshops/r-survival.html

In this example, we will work with patient data from NCCTG Lung Cancer dataset that is part of the `survival` package.

```{r imports, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(survival)
library(visR)
```

## Global Document Setup
```{r globalSetup}
# Constants
DATASET <- paste0("NCCTG Lung Cancer Dataset (from survival package ", 
                  packageVersion("survival"), ")")

# Globql formatting options
options(digits = 3)

# Global ggplot settings
theme_set(theme_classic())

# Global table settings 
options(DT.options = list(pageLength = 10, 
                          language = list(search = 'Filter:'), 
                          scrollX = TRUE))

```


## Data preparation
The variable names are changed to align with CDISC ADaM controlled terminology.

```{r initialCohort}
data(lung)

lung_cohort <- lung %>% 
    dplyr::rename(ECOG = ph.ecog,
                  Karnofsky = ph.karno,
                  SITEID = inst,
                  CNSR = status,
                  AVAL = time,
                  WEIGHT = wt.loss
                 ) %>% 
    dplyr::mutate(USUBJID = sprintf("Pat%03d", row_number()),
                  SEX = factor(if_else(sex == 1, "male", "female")),
                  CNSR = if_else(CNSR == 2, 1, 0),
                  SUBGR1 = factor(case_when(ECOG == 0 ~ "0 asymptomatic",
                                            ECOG == 1 ~ "1 ambulatory",
                                            ECOG == 2 ~ "2 in bed less than 50% of day",
                                            ECOG == 3 ~ "3 in bed more than 50% of day",
                                            ECOG == 4 ~ "4 bedbound",
                                            ECOG == 5 ~ "5 dead")
                                 ),
                  AGEGR1 = factor(case_when(age < 30 ~ "< 30y",
                                            age >= 30 & age <= 50 ~ "30-50y",
                                            age > 50 & age <= 70 ~ "51-70y",
                                            age > 70 ~ "> 70y"),
                                  levels=c("< 30y", "30-50y", "51-70y", "> 70y")
                                 )
                 ) %>%
    dplyr::select(USUBJID, SITEID, SEX, AGEGR1, WEIGHT, SUBGR1, AVAL, CNSR)
  
lung_cohort %>% 
  head(10) %>% 
  render(title = "First 10 observations of `lung_cohort`.",
                  datasource = DATASET, 
                  engine = "gt")
```

## Attrition Table
For this exercise we will only include patients with (1) ECOG available (2) non-missing weight-loss data (3) non missing censoring information and (4) positive follow-up time in our analysis. 

```{r attritionTable}

cohort_attrition <- vr_attrition_table(
  data = lung_cohort,
  criteria_descriptions = c("1. ECOG available", 
                            "2. Weight loss data available", 
                            "3. Non-missing censoring status",
                            "4. Positive follow up time"),
  criteria_conditions   = c("!is.na(SUBGR1)",
                            "!is.na(WEIGHT)",
                            "!is.na(CNSR)",
                            "AVAL >= 0"),
  subject_column_name   = 'USUBJID')

 
render(data = cohort_attrition, 
                title = "Attrition Table",
                datasource = DATASET, 
                engine = "gt")

```



## Attrition Diagram

Under construction.

```{r eval=TRUE, echo=TRUE}
complement_descriptions <- c(
"Having exclusion criterion: non ECOG",
"Having exclusion criterion: missing weight loss ", 
"Having exclusion criterion: missing censoring status", 
"Having exclusion criterion: negative follow up time")

attrition_chart_fn <- "../docs/articles/Example_analysis_files/figure-html/attrition_diagram.svg"

attrition_flow <- vr_attrition(
  N_array = cohort_attrition$`Remaining N`,
  descriptions = cohort_attrition$Criteria,
  complement_descriptions = complement_descriptions,
  display = F,
  output_path = attrition_chart_fn)
```

<div class="figure">
<img src="Example_analysis_files/figure-html/attrition_diagram.svg" alt="Attrition Diagram of Lung cohort. Data Source: NCCTG Lung Cancer Dataset (from survival package 3.2.3)"><p class="caption">Attrition Diagram of Lung cohort. Data Source: NCCTG Lung Cancer Dataset (from survival package 3.2.3)</p>
</div>

<!-- ```{r attritionFig} -->
<!-- # complement_descriptions <- c( -->
<!-- #     "Having exclusion criterion: non ECOG", -->
<!-- #     "Having exclusion criterion: missing weight loss ", -->
<!-- #     "Having exclusion criterion: missing censoring status", -->
<!-- #     "Having exclusion criterion: negative follow up time" -->
<!-- # ) -->

<!-- # Create attrition flowchart -->

<!-- #attrition_chart_fn <- "../docs/articles/Example_analysis_files/figure-html/attrition_diagram.svg" -->

<!-- #appears to be broken -->
<!-- # attrition_flow <- vr_attrition( -->
<!-- #   N_array = cohort_attrition$`Remaining N`, -->
<!-- #   descriptions = cohort_attrition$Criteria, -->
<!-- #   complement_descriptions = complement_descriptions, -->
<!-- #   output_path = file.path(tempdir(), "attrition_diagram.svg")) -->
<!-- ``` -->

<!-- ```{r attritionFigShow, fig.cap=paste("Attrition Diagram of BRCA cohort. Data Source: ", DATASET), results='asis'} -->
<!-- #knitr::include_graphics(attrition_chart_fn) -->
<!-- ``` -->

## Cohort Overview
To describe the population `tableone` can be used.
```{r table1}
# some modifications to the table for nice printing of categories
lung_cohort_tab1 <- lung_cohort %>%
  dplyr::mutate(CNSR = factor(case_when(CNSR == 0 ~ "Alive/Censored",
                                        CNSR == 1 ~ "Dead",
                                        is.na(CNSR) ~ "Missing"
                                        )
                              )
                ) %>%
  dplyr::filter(!is.na(AVAL) & !is.na(AGEGR1) & !is.na(SUBGR1)) %>%
  select(-SITEID)

# visR table1 convenience function
tableone(lung_cohort_tab1, title = "Overview over Lung Cancer patients",
             datasource = DATASET, strata = c("SEX"))
```

## Kaplan-Meier analysis
### Cohort Selection
Select cohort for the time-to-event analysis.

```{r}
lung_cohort <- lung_cohort %>%
  dplyr::filter(!is.na(SUBGR1),
                !is.na(WEIGHT),
                !is.na(CNSR),
                 AVAL >= 0
                )
```
### Kaplan Meier estimation
The function looks for variables AVAL, CNSR in the data to construct the survival formula `survival::survfit.formula(data = data, formula = Surv(AVAL, 1-CNSR) ~ strata), ...))`. This wrapper function is compatible with `%>%` and purrr::map functions without losing traceability of the dataset name.

```{r}
lung_cohort %>%
  vr_KM_est(strata = "SEX", conf.int = 0.90)
```

### Kaplan-Meier Curve
The Kaplan Meier plot is build using the pipe framework.
```{r fig.cap="Comparison of survival in male and female lung cancer patients", warning=FALSE, error = FALSE}
survfit_object <- lung_cohort %>%
  vr_KM_est(strata = "SEX", conf.int = 0.90) 

risktable <- survfit_object %>% get_risktable(min_at_risk = 0,
                statlist= c("n.risk", "n.event", "n.censor"),
                label = c("At risk", "Event", "Censor"),
                break_times = 10,
                collapse = F)
gg <-
  plot(survfit_object) %>%
  add_CI(style = "ribbon", linetype = 2) %>%
  add_CNSR(shape = 3, size = 2)

gg %>%
  add_risktable(risktable)

```

### Summary of Kaplan Meier Curve
```{r}
lung_cohort %>%
  vr_KM_est(strata = "SEX") %>%
  get_summary() %>%
  render("Overall Survival - Median survival times in days for each strata", DATASET)

lung_cohort %>%
  vr_KM_est(strata = "SEX") %>%
  get_pvalue() %>%
  render("Equality between Strata - Summary table with test of equality over strata", DATASET)
```
