---
title: Attrition Diagram 
author: Diego Saldana
date: 
output:
    bookdown::html_document2:
        code_folding: hide
        toc: true
        toc_float:
            collapsed: false
---

# visR by Example

## Attrition diagrams

A common task at the beginning of an observational study is the selection of 
a cohort by applying a set of inclusion and exclusion criteria. We will often 
want to understand how each step impacts the size of our cohort and how many 
patients are excluded at each step. This is usually shown using an attrition 
diagram. The function 'pvr_attrition' implements a simple attrition diagram, 
as shown below.

```{r results=FALSE, warning=FALSE}
library(visR)
library(knitr)

# Create figures directory 
setwd("/cloud/project/visR/notebooks")
figures_directory <- "../figures"
dir.create(figures_directory, recursive = TRUE)

# Define input to the plotting function
N_array <- c(5000, 4500, 4000, 3000)                                        
descriptions <- c(                                                              
    "All patiensts",                                                            
    "Sub-cohort 1",                                         
    "Sub-cohort 2",                                         
    "Sub-cohort 3"                                          
)                                                                               
complement_descriptions <- c(                                                   
    "Having exclusion criterion 1",                                            
    "Having exclusion criterion 2",                                            
    "Having exclusion criterion 3"                                             
)
output_path = sprintf("%s/attrition_diagram.svg", figures_directory)
cat(output_path)
pvr_attrition(N_array, descriptions, complement_descriptions, 
    output_path = output_path)
```

```{r fig.cap='An example of a simple attrition diagram'}
# Plot the attrition diagram
include_graphics("../figures/attrition_diagram.svg")
```

