---
title: "Example Survivan Analysis using visR"
author: Mark Baillie, Charlotta Fruechtenicht, Diego Saldana, Thanos Siadimas
output:
    bookdown::html_document2:
        code_folding: hide
        toc: true
        toc_float:
            collapsed: false
---

# Time to Event Analysis using visR

This short tutorial illustrates a typical use case in clinical development - the analysis of time to a certain event (e.g., death) in different populations. Typically, data obtained in randomized clinical trials (RCT) can be used to estimate the overall survival of patients in one group (e.g., treated with drug X) vs another group (e.g., treated with drug Y) and thus determine if there is a difference between these treatments.

For a more thorough introduction to Survival Analysis, we recommend the following tutorial: https://bioconnector.github.io/workshops/r-survival.html

In this example, we will work with patient data from The Cancer Genome Atlas and compare survival in primary breast cancer who received either chemotherapy or hormone therapy.

```{r installRTCGA, eval=FALSE, include=FALSE}
# install.packages("BiocManager")
# BiocManager::install()
# BiocManager::install(c("RTCGA", "RTCGA.clinical"))
```
```{r installvisR, eval=FALSE, include=FALSE}
# devtools::install_github("https://github.com/openpharma/visR.git", ref="sankey")
```


```{r imports, echo=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(RTCGA)
library(RTCGA.clinical)
library(survival)
library(visR)

library(ggsci)
library(pammtools)
library(ggpubr)
```

```{r globalSetup}
DATASET <- paste0("TCGA BRCA dataset (RTCGA version ", packageVersion("RTCGA"), ")")
```



```{r initialCohort}
brca_cohort <- survivalTCGA(BRCA.clinical, 
                     extract.cols=c("admin.disease_code", 
                                    "patient.breast_carcinoma_estrogen_receptor_status",
                                    "patient.breast_carcinoma_progesterone_receptor_status",
                                    "patient.clinical_cqcf.tumor_type",
                                    "patient.drugs.drug.therapy_types.therapy_type",
                                    "patient.age_at_initial_pathologic_diagnosis")) %>% 
  rename(tumor_type = patient.clinical_cqcf.tumor_type,
         therapy = patient.drugs.drug.therapy_types.therapy_type,
         er_status = patient.breast_carcinoma_estrogen_receptor_status,
         progesterone_status = patient.breast_carcinoma_progesterone_receptor_status,
         dx_age = patient.age_at_initial_pathologic_diagnosis,
         followup_time = times
         ) %>% 
  mutate(er_status = factor(er_status),
         progesterone_status = factor(progesterone_status),
         dx_age = as.numeric(dx_age),
         dx_age_group = factor(case_when(dx_age < 30 ~ "< 30y",
                                         dx_age >= 30 & dx_age <= 50 ~ "30-50y",
                                         dx_age > 50 & dx_age <= 70 ~ "51-70y",
                                         dx_age > 70 ~ "> 70y"),
                               levels=c("< 30y", "30-50y", "51-70y", "> 70y")))
```

# Cohort Selection
For this exercise we will only include patients with (1) primary breast cancer who have (2) received either chemotherapy or hormone therapy in our analysis

```{r attritionTable}
cohort_attrition <- vr_attrition_table(
  data = brca_cohort,
  criteria_descriptions = c("1. Primary tumors", 
                            "2. Non-missing therapy", 
                            "3. Chemotherapy or hormone therapy", 
                            "4. Non-missing vital status",
                            "5. Positive follow up time"),
  criteria_conditions   = c("tumor_type=='primary'",
                            "!is.na(therapy)",
                            "therapy %in% c('chemotherapy', 'hormone therapy')",
                            "!is.na(patient.vital_status)",
                            "followup_time >= 0"),
  subject_column_name   = 'bcr_patient_barcode'
)

brca_cohort <- brca_cohort %>% 
  filter(tumor_type=='primary',
         !is.na(therapy),
         therapy %in% c('chemotherapy', 'hormone therapy'),
         !is.na(patient.vital_status),
         followup_time >= 0)
```


```{r attritionFig, include=FALSE}
figures_directory = "../figures"

complement_descriptions <- c(                                                   
    "Having exclusion criterion: non-primary disease",                                            
    "Having exclusion criterion: missing therapy data",                                            
    "Having exclusion criterion: non chemo and non hormone therapy",
    "Having exclusion criterion: missing vital status",
    "Having exclusion criterion: negative follow up time"
)

# Create attrition flowchart
attrition_flow <- vr_attrition(cohort_attrition$`Remaining N`, cohort_attrition$Criteria, complement_descriptions,
    output_path = sprintf("%s/attrition_diagram.svg", figures_directory))
```

```{r attritionFigShow, fig.cap='Attrition Diagram of BRCA cohort', results='asis'}
# Plot the attrition diagram
knitr::include_graphics("../figures/attrition_diagram.svg")
```

```{r}
cohort_attrition %>% 
  vr_render_table("Attrition Table", "Summary of sampels fulfilling inclusion/exclusion criteria", DATASET)
```


# Cohort Overview
Prior to starting any survival analysis, we want to further inspect our cohort and see if the there are systematic differences between the chemotherapy and hormone therapy arms.
```{r table1}
brca_cohort_tab1 <- brca_cohort %>% 
  select(-bcr_patient_barcode, admin.disease_code) %>% 
  mutate(patient.vital_status = factor(case_when(patient.vital_status == 0 ~ "Alive",
                                          patient.vital_status == 1 ~ "Dead",
                                          is.na(patient.vital_status) ~ "Missing")),
         therapy = factor(therapy))

# create table
# tab1_groups <- visR::vr_create_tableone(brca_cohort_tab1, groupCols = c("therapy"))

# use alternative summary function
# tab1_groups <- visR::vr_create_tableone(brca_cohort_tab1, groupCols = c("therapy"), summary_function = visR::vr_summarize)

# render table
# tab1_groups %>% vr_render_tableone(title = "Overview over BRCA patients",
                                   # caption = "Baseline characteristics of study cohort stratified by treatment type", 
                                   # datasource = DATASET, engine = "dt")
```

```{r table1Convenience}
vr_table_one(brca_cohort_tab1, title = "Overview over BRCA patients",
             caption = "Baseline characteristics of study cohort stratified by treatment type", 
             datasource = DATASET, groupCols = c("therapy"))
```

Note: 
Number of decimal points should be a parameter
add as issue/nice to have: Order of variables: should order of variables be defined by users? selection

# Overall Survival
## Survival Object
```{r}
# currently this step is required as the function is not fully parameterized
brca_cohort_os <- brca_cohort %>% 
  rename(time = followup_time,
         status = patient.vital_status,
         trt = therapy)
```
Note: it would be good in the vr function to be able to provide user-defined names for time, event and group(s)
Also: stratification by more than one variable
Maybe pass data and formula

```{r estOS, eval=FALSE}
os_obj <- vr_est_kaplan_meier(brca_cohort_os, equation = "survival::Surv(time, status) ~ trt")
vr_plt_kaplan_meier(os_obj) +
  theme_minimal()
```
```{r estOSTable, eval=FALSE}
risk_obs <- vr_est_km_risk_table(brca_cohort_os, equation = "survival::Surv(time, status) ~ trt")
vr_plt_km_risk_table(risk_obs, time_unit = "days")
```


## Kaplan-Meier Curve
```{r fig.cap="Comparison of survival in breast cancer patients having under hormone therapy vs. chemotherapy", fig.height=6, fig.width=8, warning=FALSE}
vr_kaplan_meier(
    brca_cohort_os,
    equation = "survival::Surv(time, status) ~ trt",
    data_source = DATASET, 
    time_unit = "days",
    title = "Comparison of survival in breast cancer patients having \nunder hormone therapy vs. chemotherapy"
)
```
Nice to have: have facets with km and risk table together (maybe use patchwork?)

## Summary of Kaplan Meier Curve
```{r}
km_summary <- vr_kaplan_meier_summary(
    brca_cohort_os, "survival::Surv(time, status) ~ trt"
)

km_summary[[1]] %>% 
  vr_render_table("Overall Survival", "Median survival times for each strata", DATASET)

km_summary[[2]] %>% 
  vr_render_table("Equality between Strata", "Summary table with test of equality over strata", DATASET)
```

## Cox Proportional Hazards Model
Not quite applicable when looking at the KM curve, but we'll do it anyhow to illustrate how it works
```{r}
coxph_model <- coxph(Surv(time, status) ~ trt + er_status, data=brca_cohort_os)
```
Question: how do we best combine this with the broom object created before for exactly the same surv object (potentially)

### Show model stats

```{r}
broom::tidy(coxph_model) %>% 
  vr_render_table("Cox Proportional Hazards Model", "Estimates for all included model parameters", DATASET)
```


### Forest Plot
To Do
```{r eval=FALSE}
# TODO
```

