---
title: Attrition Diagram 
author: Diego Saldana
date: 
output:
    bookdown::html_document2:
        code_folding: hide
        toc: true
        toc_float:
            collapsed: false
---

# visR by Example

## Attrition diagrams

A common task at the beginning of an observational study is the selection of 
a cohort by applying a set of inclusion and exclusion criteria. We will often 
want to understand how each step impacts the size of our cohort and how many 
patients are excluded at each step. This is usually shown using an attrition 
diagram. The function 'pvr_attrition' implements a simple attrition diagram, 
as shown below.

```{r results=FALSE, warning=FALSE}
library(visR)
library(knitr)
library(dplyr)
library(magrittr)
library(kableExtra)

# Load data from the Veteran's Cancer Study
data(veteran)
data <- 
    veteran %>% 
    mutate(trt = as.factor(case_when(
        trt == 1 ~ "standard therapy", 
        trt == 2 ~ "test chemotherapy"
    )))

# Create Kaplan Meier estimate
broom_object <- vr_est_kaplan_meier(data)
kable(broom_object) %>% 
    kable_styling()
```

```{r results=FALSE, warning=FALSE, fig.height=8, fig.width=9}
#vr_plt_kaplan_meier(broom_object)
library(ggplot2)
library(ggsci)
library(pammtools)
library(ggpubr)


library(visR)
vr_kaplan_meier(
    data, 
    data_source = "Veteran's Lung Cancer Study", 
    title = "Comparison of survival in male patients having advanced inoperable lung cancer \nunder standard therapy vs. test chemotherapy"
)
```


```{r, fig.asp=1, fig.width=1, fig.height=1}
os_obj <- vr_est_kaplan_meier(data, equation = "survival::Surv(time, status) ~ trt")
plthex <- vr_plt_kaplan_meier(os_obj, title = "test") +
  theme_classic2() +
    theme(
    panel.background = element_rect(fill = "transparent"), # bg of the panel
    plot.background = element_rect(fill = "transparent", color = NA), # bg of the plot
    legend.position = "none",
    panel.grid.major = element_blank(), # get rid of major grid
    panel.grid.minor = element_blank(), # get rid of minor grid
    # legend.background = element_rect(fill = "transparent"), # get rid of legend bg
    # legend.box.background = element_rect(fill = "transparent") # get rid of legend panel bg
  )

ggsave("~/Downloads/KMhex.png", plot = plthex, width = 5, height = 5, units = "cm", bg = "transparent")
```

